Modelling BCG vaccination in the UK

### 2019-12-06 rmds/09-model-fitting.Rmd
model fitting corrections

Additional datasets were considered during initial model fitting and during later model calibration. These were a condensed version of the age stratified data discussed above with a reduced number of age groups (children (aged 0-14 years old), adults (15-69 years old), and older adults (70-89 years old)) and a dataset of historic pulmonary TB notifications. The advantage of condensing age groups was that the number of notifications in each group increased. This reduced the impact of stochastic noise, making fitting the model easier as trends in the data are more consistent. Secondly, reducing the number of data points, whilst still capturing the important age dynamics, reduces the compute requirements of the model (see Section \@ref(particle-filter)). As will be discussed in Section \@ref(fitting-pipeline) this was a major consideration as the model fitting approach used was highly compute intensive. The downside of this approach is that potentially important information may be lost when data is condensed into fewer groups. Historic pulmonary TB notifications (including both UK born and non-UK born cases) from 1990, 1994, and 1998 were considered as using data from the decade prior to the time period of interest allows the long term trends to be fitted to. A subset of the available data was used as this limited the impact on the compute time of the fitting pipeline. These data were originally collected in the Statutory Notifications of Infectious Diseases (NOIDS) dataset with notifications from 1913 to 1999. These data were sourced from Public Health England,[@PHE2017] and made available in `R` using `tbinenglanddataclean`^[Historic TB notification data via `tbinenglanddataclean`: https://www.samabbott.co.uk/tbinenglanddataclean/]. Downsides of using these data are that reporting standards may have changed over time so a single measurement model may not be appropriate and non-UK born cases are included in these data making fitting to this data dependent on the number of non-UK born cases pre 2000 which are themselves estimated during model fitting.

@@

```{r 09-scenarios-table}
tibble(Scenario = names(scenario_dics) %>% str_split("_") %>% 
         map_chr(~paste(., collapse = " ")), 
       DIC = unlist(transpose(scenario_dics)$dic)) %>% 
  arrange(DIC) %>% 
  mutate(Scenario = Scenario %>% 
           str_replace("plus", "and") %>% 
           str_replace("baseline", "Baseline") %>% 
           str_replace("transmission", "Transmission variable in young adults (15-29)") %>% 
           str_replace("mixing", "non-UK born mixing variable in young adults (15-29)") %>% 
  str_to_sentence()) %>% 
mutate(DIC = round(DIC, 0)) %>% 
kable(
  caption = "DIC values for each scenario evaluated during model fitting - arranged from best to worst model fit.",
  booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  column_spec(c(1), width = "10cm")
```

@@

```{r 09-age-strat-incidence, fig.scap = "Observed and predicted TB incidence stratified by model age group", fig.cap = "Observed and predicted TB incidence stratified by model age group (0-11). 0-9 refers to 5 year age groups from 0-4 years old to 45-49 years old. 10 refers to those aged between 50 and 69 and 11 refers to those aged 70+. The darker ribbon identifies the interquartile range, whilst the lighter ribbon indicates the 2.5\\% and 97.5\\% quantiles. The line represents the median."}
knitr::include_graphics(file.path(plots_path, "overview-states-1990-age-1.png"))
```

@@

The fitted model had a low effective sample size manifesting in  spuriously tight posterior distributions (Figure \@ref(fig:09-fig-prior-params); Table \@ref(tab:09-prior-posterior)). This indicates that model fitting did not allow for a full exploration of the parameter space. Given that a relatively tight proposal distribution was used this may indicate that an independent proposal is insufficient. It is likely that the key factor behind the models consistent under-prediction is the selection of a low value for the current effective contact rate and the historic effective contact rate. It is difficult to interpret these findings further due the low quality of the posterior distribution.

@@

This chapter showcased the use of `LibBi`, `rbi`, and `rbi.helpers` for fitting a complex transmission model. Unfortunately, the quality of the model fit was poor and the complexity of `LibBi` makes understanding the root cause difficult. It is possible that the model developed in the last chapter is to complex to be fitted using this approach - at least without the use of several orders of magnitude greater compute resources (or compute time) than available.[@Murray2015] However, it is also possible that even with these resources this pipeline may not have produced a high quality model fit as the model developed in the previous chapter was clearly much more complex than those envisioned by the developers of the software.[@Murray2015] Alternatively, the model itself may not be identifiable with the available data.[@Lintusaari2016] `LibBi`, `rbi` and `rbi.helpers` have great potential as a standardised toolbox for modellers. However, in their current state they are difficult to use beyond relatively simple use cases. This difficulty is compounded by sporadic, inconsistent, documentation of both the underlying software and it's `R` interface. On top of this, both `LibBi` itself and the R libraries that support it have multiple apparent bugs that can be frustrating to debug due to the layered nature of the software. `rbi` and `rbi.helpers` are under development and it is likely that many of these issues will be dealt with over time. However, `LibBi` itself has not had a major release since 2016, the community around it is largely inactive, and it's complex nature makes it difficult for newcomers to contribute towards its development. It is possible that the fitting issues outlined here may be due to errors in the code used to implement the model from the previous chapter. This is unlikely as manual prior tuning has shown that a relatively good fit to the observed data can be achieved with the current model but these results could not be reproduced using the model fitting pipeline. However, this does not mean that the modelling code is bug free. Model bugs, if present, would bias results from the fitted model rather than preventing a good fit to the data or would prevent the model from being manually tuned to fit the data.

There are many other tools available within the `R` ecosystem for fitting infectious disease models and detailing them all is beyond the scope of this thesis. However, `LibBi`, `rbi`, and `rbi.helpers` represent an attempt to provide a complete modelling framework rather than being a simple toolbox for model fitting. The `pomp` package has a similar aim making a comparison worthwhile.[@King2016] `pomp` defines models using a similar structure to that presented here and used within `LiBbi`. However, rather than using its own modelling language it relies on the use of either `R` or `C` code. This has several advantages in that `pomp` models can more easily be generalized, can be understood and implemented by users new to the package and can make use of packages from the wider `R` and `C` ecosystem. The downside to this approach is that for complex models the use of `C` is essentially a requirement for efficiency reasons and implementing complex models in `C` can be an error prone and time consuming process. `pomp` offers support for PMCMC, iterated filtering, and ABC-MCMC but does not support SMC based approaches (such as SMC-SMC and ABC-SMC).[@King2016] In developing the work presented here, and in the previous chapter, a simpler model was developed using `pomp`^[`pomp` modle code: https://gist.github.com/seabbs/f08a8a46b1342b8649df963ac015ea31]. This work was not included in this thesis as it was not sufficiently developed. However, it highlight similar issues with the `pomp` package to those observed when using `LibBi`. Whilst `pomp`'s documentation, stability, and testing were much improved over `LibBi` it had similar limitations when it came to complex models. This was to such an extent that I developed numerous helper functions to deal with both the input and output of `pomp` models^[idmodelr: https://github.com/seabbs/idmodelr]. `pomp`'s documentation also has a heavy focus on iterated filtering over other model fitting techniques.

>>>>

Additional datasets were considered during initial model fitting and during later model calibration. These were a condensed version of the age stratified data discussed above with a reduced number of age groups (children (aged 0-14 years old), adults (15-69 years old), and older adults (70-89 years old)) and a dataset of historic pulmonary TB notifications. The advantage of condensing age groups was that the number of notifications in each group increased. This reduced the impact of stochastic noise, making fitting the model easier as trends in the data are more consistent. Secondly, reducing the number of data points, whilst still capturing the important age dynamics, reduces the compute requirements of the model (see Section \@ref(particle-filter)). As will be discussed in Section \@ref(fitting-pipeline) this was a major consideration as the model fitting approach used was highly compute intensive. The downside of this approach is that potentially important information may be lost when data is condensed into fewer groups. Historic pulmonary TB notifications (including both UK born and non-UK born cases) from 1990, 1994, and 1998 were considered as using data from the decade prior to the time period of interest allows the long term trends to be fitted to. A subset of the available data was used as this limited the impact on the compute time of the fitting pipeline. These data were originally collected in the Statutory Notifications of Infectious Diseases (NOIDS) dataset with notifications from 1913 to 1999. These data were sourced from Public Health England,[@PHE2017] and made available in R using `tbinenglanddataclean`^[Historic TB notification data via `tbinenglanddataclean`: https://www.samabbott.co.uk/tbinenglanddataclean/]. Downsides of using these data are that reporting standards may have changed over time so a single measurement model may not be appropriate and non-UK born cases are included in these data making fitting to this data dependent on the number of non-UK born cases pre 2000 which are themselves estimated during model fitting.

@@

```{r 09-scenarios-table}
tibble(Scenario = names(scenario_dics) %>% str_split("_") %>% 
         map_chr(~paste(., collapse = " ")), 
       DIC = unlist(transpose(scenario_dics)$dic)) %>% 
  arrange(DIC) %>% 
  mutate(Scenario = Scenario %>% 
           str_replace("plus", "and") %>% 
           str_replace("baseline", "Baseline") %>% 
           str_replace("transmission", "Transmission variable in young adults (15-29)") %>% 
           str_replace("mixing", "non-UK born mixing variable in young adults (15-29)") %>% 
           str_to_sentence() %>% 
           str_replace("-uk", "-UK")
    ) %>% 
mutate(DIC = round(DIC, 0)) %>% 
kable(
  caption = "DIC values for each scenario evaluated during model fitting - arranged from best to worst model fit.",
  booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  column_spec(c(1), width = "10cm")
```

@@

```{r 09-age-strat-incidence, fig.scap = "Observed and predicted TB incidence stratified by model age group", fig.cap = "Observed and predicted annaul TB incidence stratified by model age group (0-11). 0-9 refers to 5 year age groups from 0-4 years old to 45-49 years old. 10 refers to those aged between 50 and 69 and 11 refers to those aged 70+. The darker ribbon identifies the interquartile range, whilst the lighter ribbon indicates the 2.5\\% and 97.5\\% quantiles. The line represents the median."}
knitr::include_graphics(file.path(plots_path, "overview-states-1990-age-1.png"))
```

@@

The fitted model had a low effective sample size manifesting in  spuriously tight posterior distributions (Figure \@ref(fig:09-fig-prior-params); Table \@ref(tab:09-prior-posterior)). This indicates that model fitting did not allow for a full exploration of the parameter space. Given that a relatively tight proposal distribution was used this may indicate that an independent proposal is insufficient. It is likely that the key factor behind the model's consistent under-prediction is the selection of a low value for the current effective contact rate and the historic effective contact rate. It is difficult to interpret these findings further due the low quality of the posterior distribution.

@@

This chapter showcased the use of `LibBi`, `rbi`, and `rbi.helpers` for fitting a complex transmission model. Unfortunately, the quality of the model fit was poor and the complexity of `LibBi` makes understanding the root cause difficult. It is possible that the model developed in the last chapter is to complex to be fitted using this approach - at least without the use of several orders of magnitude greater compute resources (or compute time) than available.[@Murray2015] However, it is also possible that even with these resources this pipeline may not have produced a high quality model fit as the model developed in the previous chapter was clearly much more complex than those envisioned by the developers of the software.[@Murray2015] Alternatively, the model itself may not be identifiable with the available data.[@Lintusaari2016] `LibBi`, `rbi` and `rbi.helpers` have great potential as a standardised toolbox for modellers. However, in their current state they are difficult to use beyond relatively simple use cases. This difficulty is compounded by sporadic, inconsistent, documentation of both the underlying software and it's R interface. On top of this, both `LibBi` itself and the R libraries that support it have multiple apparent bugs that can be frustrating to debug due to the layered nature of the software. `rbi` and `rbi.helpers` are under development and it is likely that many of these issues will be dealt with over time. However, `LibBi` itself has not had a major release since 2016, the community around it is largely inactive, and it's complex nature makes it difficult for newcomers to contribute towards its development. It is possible that the fitting issues outlined here may be due to errors in the code used to implement the model from the previous chapter. This is unlikely as manual prior tuning has shown that a relatively good fit to the observed data can be achieved with the current model but these results could not be reproduced using the model fitting pipeline. However, this does not mean that the modelling code is bug free. Model bugs, if present, would bias results from the fitted model rather than preventing a good fit to the data or would prevent the model from being manually tuned to fit the data.

There are many other tools available within the R ecosystem for fitting infectious disease models and detailing them all is beyond the scope of this thesis. However, `LibBi`, `rbi`, and `rbi.helpers` represent an attempt to provide a complete modelling framework rather than being a simple toolbox for model fitting. The `pomp` package has a similar aim making a comparison worthwhile.[@King2016] `pomp` defines models using a similar structure to that presented here and used within `LiBbi`. However, rather than using its own modelling language it relies on the use of either R or C code. This has several advantages in that `pomp` models can more easily be generalized, can be understood and implemented by users new to the package and can make use of packages from the wider R and C ecosystem. The downside to this approach is that for complex models the use of C is essentially a requirement for efficiency reasons and implementing complex models in C can be an error prone and time consuming process. `pomp` offers support for PMCMC, iterated filtering, and ABC-MCMC but does not support SMC based approaches (such as SMC-SMC and ABC-SMC).[@King2016] In developing the work presented here, and in the previous chapter, a simpler model was developed using `pomp`^[`pomp` modle code: https://gist.github.com/seabbs/f08a8a46b1342b8649df963ac015ea31]. This work was not included in this thesis as it was not sufficiently developed. However, it highlight similar issues with the `pomp` package to those observed when using `LibBi`. Whilst `pomp`'s documentation, stability, and testing were much improved over `LibBi` it had similar limitations when it came to complex models. This was to such an extent that I developed numerous helper functions to deal with both the input and output of `pomp` models^[idmodelr: https://github.com/seabbs/idmodelr]. `pomp`'s documentation also has a heavy focus on iterated filtering over other model fitting techniques.
